---
title: "Race Results Analysis"
author: "Elaine McVey"
date: "January 14, 2016"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}

library(knitr)

knitr::opts_chunk$set(echo = TRUE)

knit_hooks$set(inline = function(x) {
  if (is.numeric(x)) {
    prettyNum(x, big.mark = ",")
  } else {
    x
  }
})

```

This analysis summarizes and compares race results by age and gender.


## Raw Data

The raw data used in this analysis is a set of individual race results including finish time, pace, age, and gender for each athlete.  Bib number is the unique identifier, and event name is a free text description of the race.

```{r read_data}

# Load the required packages
library(readr)

# Read in the raw race results
results_df <- read_csv('output/results_spartan_sc_sat_beast_open.csv')

# Show the top of the dataframe
kable(head(results_df))

```

## Calculations

To compare across age and gender, ranks and percentiles within categories and age groups must be calculated.

```{r calc_vars, message = FALSE}

# Load the required packages
library(dplyr)

# Source the age group assignment function
source('R/group_age.R')

# Create the age groups
results_df <- results_df %>%
  mutate(age_cat = factor(group_age(age), 
                          levels = c('under 18', '18-24', '25-29', '30-34', '35-39', '40-44', 
                                     '45-49', '50-54', '55-59', '60-64', '65 and up')), # categorize age
         age_group = paste0(gender, '_', age_cat)) %>% # combine age category with gender for age groups
  filter(!is.na(age_cat)) 

# Calculate ranks within subgroups
results_df <- results_df %>%
  arrange(finish_time) %>%
  mutate(rank_overall = 1:n()) %>%
  group_by(age_group) %>%
  mutate(rank_agegrp = 1:n()) %>%
  group_by(gender) %>%
  mutate(rank_gender = 1:n()) %>%
  ungroup()
    
# Calculate percentages within subgroups
results_df <- results_df %>%
  arrange(finish_time) %>%
  mutate(pct_overall = round(100 * rank_overall/max(rank_overall), 2)) %>%
  group_by(age_group) %>%
  mutate(pct_agegrp = round(100 * rank_agegrp/max(rank_agegrp), 2)) %>%
  group_by(gender) %>%
  mutate(pct_gender = round(100 * rank_gender/max(rank_gender), 2)) %>%
  ungroup %>%
  arrange(finish_time)

# Show the top of the dataframe
kable(head(results_df %>% select(-event_name, -bib_number)))

```

## Graphical Analysis

### Who participated

A total of `r sum(results_df$gender == 'F')` women and `r sum(results_df$gender == 'M')` men participated in the `r unique(results_df$event_name)`.

```{r demo_hist}

library(ggplot2)

# Count of participants by age group
results_df %>%
  ggplot(aes(x = age_cat)) +
  geom_bar() +
  theme_bw() +
  facet_wrap( ~ gender) +
  xlab('') +
  theme(axis.text.x = element_text(angle = 90))
  
```

### Pace by gender

```{r cdist_pace_gender, message = FALSE}

results_df %>%
  filter(pct_gender <= 95) %>%
  ggplot(aes(x = pace, y = pct_gender, color = gender)) + 
  geom_line() +
  theme_bw() +
  xlab('Pace (min/mile)') + ylab('Percentile within gender') +
  ggtitle('Cumulative Pace Distribution by Gender') 

```

### Pace by age 

```{r scatter_pace_age, message = FALSE}

results_df %>%
  filter(pct_agegrp <= 95) %>%
  ggplot(aes(x = age, y = pace, color = gender)) + 
  geom_point(alpha = .5) +
  geom_smooth() +
  facet_wrap( ~ gender, ncol = 1) +
  theme_bw() +
  ylab('Pace (min/mile)') + xlab('Age') 

```

### Gender gap

Adjusting for gender, how do athletes in the same performance percentiles of their gender compare on pace?  Is the gender gap consistent across the performance range?

```{r gender_gap, message = FALSE, warning = FALSE}

library(tidyr)

# Calculate the mean pace within each percentile of gender
gap_df <- results_df %>%
  mutate(pct_gender = ceiling(pct_gender)) %>%
  group_by(gender, pct_gender) %>%
  summarize(mean_pace = mean(pace))

# Reorganize to calculate the difference in mean pace between genders at each percentile
gap_wide_df <- gap_df %>%
  spread(gender, mean_pace) %>%
  mutate(gap_gender = F - M) 

kable(head(gap_wide_df))

```

```{r plot_gap, message = FALSE, warning = FALSE}

gap_wide_df %>%
  filter(pct_gender <= 80) %>%
  ggplot(aes(x = pct_gender, y = gap_gender)) + theme_bw() +
  geom_line() + 
  expand_limits(y = 0) +
  xlab('Percentile within gender') +
  ylab('Gender gap (minutes/mile)') 

```

```{r}
save(results_df, gap_wide_df, file = 'output/spartan_processed.rda')
```

